TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-09 src/paper.tex
minor_polish_first

We use the pretrained self-supervised learning (SSL) model WavLM Large \cite{wavlm} to encode speech into continuous representations. Specifically, we extract the outputs from six hidden layers: 1, 3, 7, 12, 18, and 23. Given a speech signal \(s \in \mathbb{R}^{T'}\), the output of WavLM is a tensor \(\bm{r}\) with shape \(n \times T \times E\), where \(n\) is the number of output layers (6 in this case), \(T\) is the time dimension, and \(E\) represents the embedding dimension, which is 1024 in WavLM Large.  For tokenization, we apply separate Kmeans models to each output layer, with each model using the same number of clusters, denoted by \(K\). After tokenization, the continuous embedding \(\bm{r}\) is transformed into a discrete tensor \(\bm{d}\) with shape \(n \times T\), where each value \(\bm{d}_{i} \in (0, K-1) \). In all our experiments, we set \(K = 1000\). For both reference and mixed speech, the same Kmeans model and the same layer combination from WavLM Large are used. The encoder remains frozen during training.
The encoding strategy for mixed speech is crucial to the performance of the model. Given a reference speech \(s_r \in \mathbb{R}^{T^r}\) and a mixed speech \(s_m \in \mathbb{R}^{T'}\), we follow the previously described procedure to encode the reference speech into a tensor \(\bm{d_r}\) of shape \(n \times T_r\). However, for mixed speech, instead of applying the encoding directly, we first concatenate it with the reference speech, creating a signal \(s' = [s_r, s_m, s_r] \in \mathbb{R}^{(T^r + T' + T^r)}\). This concatenated signal is then input into the encoder, producing an output tensor \(\bm{d'}\) with shape \(n \times (T_r + T + T_r)\), where \(T\) represents the time dimension for the mixed speech without concatenation. The tensor \(\bm{d'}\) contains discrete tokens for the two segments of reference speech and the mixed speech. We extract the portion \(\bm{d}\) corresponding only to the mixed speech, resulting in an output tensor of shape \(n \times T\).
This approach is inspired by the training schema of WavLM \cite{wavlm}, which trains the model by overlapping clean speech with an interfering signal that covers less than 50\% of the clean speech's duration, using the first utterance as the primary speaker. This allows WavLM to identify and focus on the main speaker when producing embeddings. Our experiments demonstrated that this concatenation strategy significantly enhances the model's performance by guiding it to prioritize the target speaker's information.
\subsection{Modeling}
\subsubsection{Attention Embedding}
After obtaining the discrete tensor \(\bm{d}\) with shape \(6 \times T\), we use 6
learnable embedding tables each with \(K\) entires to embed the 6 layers 
respectively, each resulting in a tensor of shape \(T \times E\). 
After embedding, we follow the same recipe as in \cite{dasb} to 
aggregate the tensor by using attention mechanism to sum all the 6 tensors. This summation 
keeps the information of each layer while reducing the system complexity by reducing the 
dimension of layers. After attention embedding, we obtain reference embedding \(E_r\) and 
mixture embedding \(E_m\).

@@

We use the \texttt{dev} set of Libri2Mix \cite{librimix} to test our model performance and compare it with 
other baseline models.

>>>>

We use the pretrained self-supervised learning (SSL) model WavLM Large \cite{wavlm} to encode speech into continuous representations. Specifically, we extract the outputs from six hidden layers: 1, 3, 7, 12, 18, and 23. Given a speech signal \(s \in \mathbb{R}^{T'}\), the output of WavLM is a tensor \(\bm{r}\) with shape \(n \times T \times E\), where \(n\) is the number of output layers (6 in this case), \(T\) is the time dimension, and \(E\) represents the embedding dimension, which is 1024 in WavLM Large.  For tokenization, we apply separate Kmeans models to each output layer, with each model using the same number of clusters, denoted by \(K\). After tokenization, the continuous embedding \(\bm{r}\) is transformed into a discrete tensor \(\bm{d}\) with shape \(n \times T\), where each value \(\bm{d}_{i} \in (0, K-1) \). In all our experiments, we set \(K = 1000\). For both reference and mixed speech, the same Kmeans model and the same layer combination from WavLM Large are used. The encoder remains frozen during training.
The encoding strategy for mixed speech is crucial to the performance of the model. Given a reference speech \(s_r \in \mathbb{R}^{T^r}\) and a mixed speech \(s_m \in \mathbb{R}^{T'}\), we follow the previously described procedure to encode the reference speech into a tensor \(\bm{d_r}\) of shape \(n \times T_r\). However, for mixed speech, instead of applying the encoding directly, we first concatenate it with the reference speech, creating a signal \(s' = [s_r, s_m, s_r] \in \mathbb{R}^{(T^r + T' + T^r)}\). This concatenated signal is then input into the encoder, producing an output tensor \(\bm{d'}\) with shape \(n \times (T_r + T + T_r)\), where \(T\) represents the length for the mixed speech embedding. The tensor \(\bm{d'}\) contains discrete tokens for the two segments of reference speech and the mixed speech. We extract the portion \(\bm{d}\) corresponding to the mixed speech, resulting in an output tensor of shape \(n \times T\).

This approach is inspired by WavLM \cite{wavlm}, which trains the model by overlapping clean speech with an interfering signal covering less than 50\% of the speech, using the first utterance as the primary speaker. This allows WavLM to focus on producing target speaker dependent embeddings. Our experiments demonstrated that this concatenation strategy significantly enhances the model's performance by guiding it to prioritize the target speaker's information.
\subsection{Modeling}
\subsubsection{Attention Embedding}
After obtaining the discrete tensor \(\bm{d}\) with shape \(6 \times T\), we use 6
learnable embedding tables each with \(K\) entires to embed the 6 layers 
respectively, each resulting in a tensor of shape \(T \times E\). 
After embedding, we follow the same recipe as in \cite{dasb} to 
aggregate the tensor by using attention mechanism to sum all the 6 tensors. This summation 
keeps the information of each layer while reducing the system complexity by reducing the 
dimension of layers. After attention embedding, we obtain reference embedding \(E_r\) and 
mixture embedding \(E_m\).

@@

We use the \texttt{test} set of Libri2Mix \cite{librimix} to test our model performance and compare it with 
other baseline models.
